---
title: "scRNA Analysis - 04 - Integrated Dataset Analysis"
author: "Sydney Informatics Hub"
format: html
---

This notebook is designed to run downstream analyses on an integrated Seurat dataset. This includes automatic and (optionally) manual cell type annotation, cell cycle annotation, differential gene expression analysis, and pathway analysis.

## Imports

While working through this notebook manually, you will need to run the following block of code to import all the necessary R libraries and helper functions:

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(Seurat)
library(scuttle)
library(SingleR)
source("R/helpers.R")
source("R/shinyapps.R")
```

## Read in dataset

```{r inputs}
dataset <- "outputs/integrated_dataset.filtered_clustered.doublets_detected.integrated.Rds"

# Load data
integrated <- readRDS(dataset)
```

## Annotation 1: Cell Cycle

Cell cycle annotation is a common step, particularly for cancer data sets.

```{r cell_cycle_scoring}
gene2ens <- integrated@assays$RNA@meta.data
s.genes.ens <- gene2ens$gene_versions[match(cc.genes$s.genes, gene2ens$gene_symbols)]
s.genes.ens <- s.genes.ens[!is.na(s.genes.ens)]
g2m.genes.ens <- gene2ens$gene_versions[match(cc.genes$g2m.genes, gene2ens$gene_symbols)]
g2m.genes.ens <- g2m.genes.ens[!is.na(g2m.genes.ens)]

integrated <- CellCycleScoring(
  integrated,
  s.features = s.genes.ens,
  g2m.features = g2m.genes.ens
)
```

### Plot cell cycle annotations

```{r plot_cell_cycle_scoring}
DimPlot(integrated, reduction = "umap", group.by = "Phase")
```

## Annotation 2: Automatic cell type assignment with HPCA

This step will assign cell types to each cell using the `SingleR` and `celldex` packages and the HPCA database.

```{r annotate_cell_types_hpca}
ref <- celldex::HumanPrimaryCellAtlasData()

sce <- as.SingleCellExperiment(integrated, assay = "RNA")
sceM <- logNormCounts(sce)

# Map Ensembl IDs to gene IDs
ens2gene <- AnnotationDbi::mapIds(
  EnsDb.Hsapiens.v86::EnsDb.Hsapiens.v86,
  keys = rownames(sce),
  column = "SYMBOL",
  keytype = "GENEID"
)
stopifnot(all(rownames(sceM) == names(ens2gene)))
keep <- !is.na(ens2gene)
ens2gene <- ens2gene[keep]
sceM <- sceM[keep,]
rownames(sceM) <- ens2gene

predicted <- SingleR(test = sceM, ref = ref, labels = ref$label.main)
keep <- table(predicted$labels) > 10  # NOTE: Should we change this?
integrated$SingleR.hpca_main <- ifelse(keep[predicted$labels], predicted$labels, "Other")

rm(sce, sceM, ens2gene, keep, predicted, ref)
```

### Plot cell type assignment

```{r plot_cell_types_hpca}
DimPlot(
  integrated,
  reduction = "umap",
  group.by = c(cluster_name, "SingleR.hpca_main"),
  label = TRUE,
  repel = TRUE,
  label.box = TRUE
)
```

### Print table of cluster assignments

```{r print_cluster_assignments}
integrated@meta.data %>%
  dplyr::select(any_of(cluster_name), SingleR.hpca_main) %>%
  group_by(.data[[cluster_name]], SingleR.hpca_main) %>%
  summarise(n_cells = n()) %>%
  pivot_wider(names_from = cluster_name, values_from = n_cells)
```

## Annotation 3: Manual cluster-level annotation with curated gene lists

If you have a list of cell types with curated cell markers available, you can supply a TSV file defining these gene programs. The TSV file should contain two columns with the headers `cell_type` and `gene_ids`. The `gene_ids` column should contain all of the marker genes associated with the cell type, separated by commas. For example:

```
cell_type gene_ids
T_NK  CD3D,CD3E,CD3G,CD247
B_cells CD79A,CD79B,IGHM,IGHD
Myeloid LYZ,S100A8,S100A9,CD14
Fibroblasts COL1A1,COL1A2,COL3A1,ACTA2
Endothelial PECAM1,CD34,CDH5,VWF
```

```{r read_custom_programs}
custom_marker_genes_file <- "data/custom_markers.tsv"
custom_marker_genes <- read_tsv(custom_marker_genes_file)

custom_programs <- custom_marker_genes$cell_type
custom_marker_genes <- strsplit(custom_marker_genes$gene_ids, ",")
names(custom_marker_genes) <- custom_programs

custom_marker_genes
```

```{r get_ensembl_gene_ids}
custom_marker_genes_ens <- lapply(custom_marker_genes, function(g) { gene2ens$gene_versions[match(g, gene2ens$gene_symbols)] })
custom_marker_genes_ens
```

```{r score_cells}
integrated <- AddModuleScore(
  integrated,
  features = custom_marker_genes_ens,
  name = names(custom_marker_genes_ens)
)

# Remove numeric suffix from cluster names
numeric_suffix_clusters <- paste0(names(custom_marker_genes_ens), 1:length(custom_marker_genes_ens))
stopifnot(all(numeric_suffix_clusters %in% colnames(integrated@meta.data)))  # Sanity check
colnames(integrated@meta.data)[colnames(integrated@meta.data) %in% numeric_suffix_clusters] <- names(custom_marker_genes_ens)
stopifnot(all(names(custom_marker_genes_ens) %in% colnames(integrated@meta.data)))  # Sanity check
```

```{r get_max_scores}

```

## Write data to file

```{r save_data, eval = FALSE}
dir.create("outputs")
SaveSeuratRds(integrated, "outputs/integrated_dataset.filtered_clustered.doublets_detected.integrated.analysed.Rds")
```
